---
title: |
  [1]: https://www.facebook.com/An%C3%A1lisis-y-visualizaci%C3%B3n-de-datos-100602148375744
  [2]: https://raw.githubusercontent.com/DataFeast71/COVID19_plots/main/img/Logo_W.jpeg {width=1in}
  [![Analisis Y visualizacion][2]][1]
  Food Loss
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: null
    df_print: paged
    highlight: zenburn
    theme: cerulean
    toc: true
    toc_float: true
    toc_deep: 3
---

<style>
.tocify-extend-page {
  height: 0 !important;
}
</style>

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(networkD3)
```

## Data

```{r}
food_raw <- read.csv("Dataset/Data_All.csv", header = TRUE, stringsAsFactors = FALSE)
head(food_raw, 2)
```

```{r}
dim(food_raw)
```

The data set contains 18754 rows and 22 columns

## Data Cleaning and Preparation

Numbers of NaN values.

The columns treatments and units can be removed. The problem that we see is that the empty cells aren't `NaN` so we need to replace that or remove it.

```{r}
apply(food_raw, 2, function(x) sum(is.na(x)))
```

In this case the column region, loss_monetary, loss_qualitiative, period of storage can be removed.

```{r}
apply(food_raw, 2, function(x) sum(x == ""))
```

Now we want to know how many unique values are in each column:

```{r}
apply(food_raw, 2, function(x) length(unique(x)))
```

The variables the could be importan for us are:

* **Crop** which have 123 unique values

One way that we found to classify the crops is using the basket group data from the webpage. In this they make subsets for each basket class so we can use that to classify.

```{r}
basket <- read.csv("Dataset/Crops_Basket.csv", header = TRUE, stringsAsFactors = FALSE)
```

* **fsc_location1**, This column contains the supply chain and we note that there are some mistakes in Harvest, Traders.

```{r}
unique(food_raw$fsc_location1)
```

We would correct that and select some important supply chains.

```{r}
# typo correction in fsc_location1
food_raw$fsc_location1 <- ifelse(food_raw$fsc_location1 == "Havest","Harvest",food_raw$fsc_location1) 
food_raw$fsc_location1 <- ifelse(food_raw$fsc_location1 == "Trader","Traders",food_raw$fsc_location1) 
```

```{r}
supplychain <- c("Pre-Harvest","Harvest","Grading","Storage",
                 "Transport","Traders","Processing","Packaging",
                 "Distribution","Wholesale","Retail","Consumer",
                 "Export", "Farm", "Restaurants", "Market")

food_raw <- food_raw %>% 
  filter(fsc_location1 %in% supplychain)
```

## Data seleccion for Sankey

### Crops and countries

For the moment we would select only the data from 2017 and the variable that is importan for us is `loss_per_clean`

```{r}
food_2017 <- food_raw %>% 
  filter(timepointyears == 2017) %>% 
  select(country, crop, loss_per_clean, fsc_location1) %>%
  left_join(basket, by = "crop") %>% 
  distinct()
head(food_2017,2)
```

For the Sankey diagram we want to show the basket food class then which crops are present and finally the country related to the food loss.

```{r, message=FALSE, warning=FALSE}
food_loss <- food_2017 %>% 
  select(Class, crop, country, loss_per_clean) %>% 
  group_by(Class, crop, country) %>% 
  summarise(loss = sum(loss_per_clean))
food_loss[1:5,]
```

```{r, message=FALSE, warning=FALSE}
links <- food_loss %>% 
  group_by(Class, crop) %>% 
  summarise(loss = sum(loss)) %>% 
  ungroup() %>% 
  rename(source = Class, target = crop)

links2 <- food_loss %>% 
  group_by(crop, country) %>% 
  summarise(loss = sum(loss)) %>% 
  ungroup() %>% 
  rename(source = crop, target = country)

links <- rbind(links, links2)
```

```{r}
# From these flows we need to create a node data frame: it lists every entities involved in the flow
nodes <- data.frame(
  name=c(as.character(links$source), 
  as.character(links$target)) %>% unique()
)
```

```{r}
# With networkD3, connection must be provided using id, not using real name like in the links dataframe.. So we need to reformat it.
links$IDsource <- match(links$source, nodes$name)-1 
links$IDtarget <- match(links$target, nodes$name)-1
```

```{r}
# Make the Network
p <- sankeyNetwork(Links = links, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              Value = "loss", NodeID = "name", 
              sinksRight=FALSE)
p
```

### Class, supply country

In this case we want to know the food loss by the supply chain.

```{r}
food_2017 <- food_raw %>% 
  filter(timepointyears == 2017) %>% 
  select(country, crop, loss_per_clean, fsc_location1) %>%
  left_join(basket, by = "crop") %>% 
  distinct()
head(food_2017,2)
```

For the Sankey diagram we want to show the basket food class then which crops are present and finally the country related to the food loss
```{r, message=FALSE, warning=FALSE}
food_loss <- food_2017 %>% 
  select(fsc_location1, Class, country, loss_per_clean) %>% 
  group_by(fsc_location1, Class, country) %>% 
  summarise(loss = sum(loss_per_clean))
```

```{r, message=FALSE, warning=FALSE}
links <- food_loss %>% 
  group_by(Class, fsc_location1) %>% 
  summarise(loss = sum(loss)) %>% 
  ungroup() %>% 
  rename(source = Class, target = fsc_location1)

links2 <- food_loss %>% 
  group_by(fsc_location1, country) %>% 
  summarise(loss = sum(loss)) %>% 
  ungroup() %>% 
  rename(source = fsc_location1, target = country)

links <- rbind(links, links2)
```

```{r}
# From these flows we need to create a node data frame: it lists every entities involved in the flow
nodes <- data.frame(
  name=c(as.character(links$source), 
  as.character(links$target)) %>% unique()
)
```

```{r}
# With networkD3, connection must be provided using id, not using real name like in the links dataframe.. So we need to reformat it.
links$IDsource <- match(links$source, nodes$name)-1 
links$IDtarget <- match(links$target, nodes$name)-1
```

```{r}
# Make the Network
p <- sankeyNetwork(Links = links, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              Value = "loss", NodeID = "name", 
              sinksRight=FALSE)
p
```

### Sankey plot by country 

Another Sankey plot that we can do is selecting only one country and see how it is the behavior in a specific year. For example, select Mexico in 2016.

```{r}
food_country <- food_raw %>% 
  filter(timepointyears == 2016 & country == "Mexico") %>% 
  select(country, crop, loss_per_clean, fsc_location1) %>%
  left_join(basket, by = "crop") %>% 
  distinct() %>% 
  rename(loss = loss_per_clean)
```

To make easy the process we would try to make a function to prepare the data for Sankey.

```{r}
sankey_data <- function(df, link1, link2) {
  # Make the links
  links1 <- df %>% 
  group_by(Class, get(link1)) %>% 
  summarise(loss = sum(loss)) %>% 
  ungroup() 
  names(links1) <- c("source", "target","loss")

  links2 <- df %>% 
  group_by(get(link1), get(link2)) %>% 
  summarise(loss = sum(loss)) %>% 
  ungroup() 

  names(links2) <- c("source", "target","loss")
  
  links <- rbind(links1, links2)
  
  # From these flows we need to create a node data frame: it lists every entities involved in the flow
  nodes <- data.frame(
    name = c(as.character(links$source),
             as.character(links$target)) %>% unique()
    )
  
  # With networkD3, connection must be provided using id, not using real name like in the links dataframe.. So we need to reformat it.
  links$IDsource <- match(links$source, nodes$name)-1 
  links$IDtarget <- match(links$target, nodes$name)-1
  
  return(list(links = links, nodes = nodes))
}
```

```{r}
network <- sankey_data(food_country, link1 = "fsc_location1", link2 = "crop")
```

```{r}
# Make the Network
p <- sankeyNetwork(Links = network$links, Nodes = network$nodes,
              Source = "IDsource", Target = "IDtarget",
              Value = "loss", NodeID = "name", 
              sinksRight=FALSE, fontSize = 10, units = "%")
p
```

Cool!!

```{r}
food_raw %>% 
  select(country, crop, loss_per_clean, fsc_location1, timepointyears) %>%
  left_join(basket, by = "crop") %>% 
write.csv("Data_app.csv", row.names = FALSE)
```

